---
title: "Exploratory Data Analysis Using R"
output:
  html_document:
    df_print: paged
---

```{r}
library(tidyverse)
```

```{r}
# Load the Motor Trend Car Road Tests (mtcars) dataset
carData = read.csv("C:/Users/User/Downloads/MAHE FILES/MAHE SEMESTER 1/R and Python/MAHE R WK3&4/mtcars3.csv", header = TRUE)
head(carData, 5)
```
```{r}
# Create a vector of categorical columns
categorical_cols = c('vs', 'am')

# Convert the columns to factor type
carData[categorical_cols] = lapply(carData[categorical_cols], as.factor)
```

```{r}
# Add a new column called cyltype with value High
# is cyl is greater than 4 and Low otherwise
carData = carData %>% mutate(cyltype = ifelse(cyl > 4, 'High', 'Low'))
head(carData)
```

```{r}
# Summarize the features
summary(carData)
```

```{r}
# Visualize distribution of a categorical
# variable using bar chart
ggplot(data = carData) +
  geom_bar(aes(x = cyltype))
```

```{r}
# Count the number of observations in each category
carData %>% count(cyltype)
```

```{r}
# Visualize distribution of a continuous
# variable using histogram
ggplot(data = carData) +
  geom_histogram(aes(x = mpg), binwidth = 2.5)
```

```{r}
# Visualzing the histogram using counts
#cut_width is a ggplot fxn for generating numerical values of a histogram
#cut_width (2.5) basically splits the range of variable-mpg values into blocks that are of 2.5 units apart. And then count fxn will count how many falls into each block.
#In summary, cut_width acts like class-interval
carData %>% 
  count(cut_width(mpg, 2.5))
```

```{r}
# Visualizing multiple histograms
#Sometime, creating a multiple barchart on one chart might cause overlapping of the boxes. Therefore, we use the straight line called freq polygon 
#freqpoly is a Frequency Polygon object
ggplot(data = carData, mapping = aes(x = mpg)) +
  geom_freqpoly(binwidth = 2.5, mapping = aes(colour = cyltype))
```
The red line ("High") shows a peak in frequency around the lower mpg range (approximately 15-20 mpg), indicating vehicles in this category are less fuel-efficient.
The blue line ("Low") peaks at higher mpg ranges (approximately 25-30 mpg), suggesting that these vehicles are more fuel-efficient.
The data for the two groups have little overlap, with distinct trends in different ranges of mpg.

```{r}
# Boxplot to visualize the covariance
# between a continuous and categorical
#To show the distribution of a categorical variable over a continous variable
# feature
ggplot(data = carData, mapping = aes(x = cyltype, y = mpg)) +
  geom_boxplot()
```

```{r}
# Load the mpg dataset
data('mpg')
mpgData = mpg
head(mpgData)
```

```{r}
# Boxplot to visualize highway mpg according to
# car type
ggplot(data = mpgData, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()
```

```{r}
# Reorder boxplot according to median 
#I want to rearrange the boxplot according to their median. Then i use the reorder fxn. That is, reorder the 'class' var using the values of thr 'hwy' var wrt their median.
# to visualize the trend
ggplot(data = mpgData, mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  geom_boxplot()
```

```{r}
# Flip the boxplot for better visualization
#coord_flip changes the position of the axes
ggplot(data = mpgData) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), y = hwy)) +
  coord_flip()
```

```{r}
# Visualize covariance between two
# categorical features
carData %>% count(cyltype, am)
ggplot(data = carData) +
  geom_count(mapping = aes(x = cyltype, y = am))
```

```{r}
# Visualize covariance between two
# continuous features - create a
# scatter plot of mpg vs. HP
ggplot(data = carData, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point(size = 1) 
```

```{r}
# Load the diamonds dataset
data(diamonds)
head(diamonds)
```

```{r}
# Visualize covariance between two
# continuous features - create a
# scatter plot of carat vs. price
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

```{r}
# Load the hexbin package
library(hexbin)
```

```{r}
ggplot(data = diamonds) +
  geom_hex(mapping = aes(x = carat, y = price))
```




