---
title: "Data Visualization using ggplot2 and dplyr in R"
author: "Ezebuike Esther"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
#Import packages & dataset 
library(dplyr)
library(ggplot2)
gapminder <- read.csv("C:/Users/User/Downloads/GAPMINDER DATASET/gapminder_full.csv", header = TRUE, sep = ",")
```


```{r}
head(gapminder,10)
```


```{r}
#To create a subset of the gapminder dataset with the year=1957
gapminder_1957 <- gapminder %>% filter(year == 1957)
```

```{r}
#Plot a scatter plot of the population on the x-axis and the life expectancy on the y-axis

ggplot(gapminder_1957, aes(x = population, y = life_exp)) +    geom_point()
```


```{r}
#Plot a scatter plot of the population on the x-axis and the gdp per capita on the y-axis

ggplot(gapminder_1957, aes(x = population, y = gdp_cap)) +    geom_point()
```


```{r}
#Plot a scatter plot of the gdp per capita on the x-axis and the life expectancy on the y-axis

ggplot(gapminder_1957, aes(x = gdp_cap, y = life_exp)) +    geom_point()
```


```{r}
#Adding log scale
#Plot a scatter plot of the population on the y-axis and the life expectancy on the y-axis

ggplot(gapminder_1957, aes(x = population, y = life_exp)) +    geom_point() + scale_x_log10()
```
Polpulation increases as life expectancy increases.


```{r}
#Plot a scatter plot of the population on the y-axis and the gdp per capita on the y-axis

ggplot(gapminder_1957, aes(x = population, y = gdp_cap)) +    geom_point() + scale_x_log10() + scale_y_log10()
```
Population rises with an increase in gdp/capita

```{r}
#Using the color and size aesthetic
#Plot a scatter plot of the population on the y-axis and the life expectancy on the y-axis with color representing the continent

ggplot(gapminder_1957, aes(x = population, y = life_exp, color = continent)) + 
  geom_point() + scale_x_log10()
```
Africans have lower life expectancy. Europeans have higher life expectancy.


```{r}
#Using the color and size aesthetic
#Plot a scatter plot of the population on the y-axis and the life expectancy on the y-axis with color representing the continent and size representing the gdp per capita

ggplot(gapminder_1957, aes(x = population, y = life_exp, color = continent, size = gdp_cap)) + 
  geom_point() + scale_x_log10()
```
Africans had lower life expectancy with smaller gdp_cap. Asians had bigger gdp_cap and a moderate life expectancy.


```{r}
#Using faceting
#Plot a scatter plot comparing population on the x-axis and the life expectancy on the y-axis facetted by continent

ggplot(gapminder_1957, aes(x = population, y = life_exp)) +    geom_point() + scale_x_log10() + facet_wrap(~continent)
```


```{r}
#Plot a scatter plot of the gdp per capita on the x-axis and the life expectancy on the y-axis with color represented by continent, size represented by population and facetted by year

ggplot(gapminder, aes(x = gdp_cap, y = life_exp, color = continent, size = population)) + 
  geom_point() + scale_x_log10() + facet_wrap(~year)
```
Life expectancy and gdp_cap increase globally over the years. Asians tend to have bigger gdp_cap.


```{r}
#VISUALIZING WITH SUMMARIZED DATA
#creating a variable by_year that get the median life expectancy for each year
by_year <- gapminder %>% group_by(year) %>% summarise(medianLifeExp = median(life_exp))
head(by_year, 10)
```


```{r}
#create a scatter plot to show the change in median life expectancy using the by_year data

ggplot(by_year, aes(x = year, y = medianLifeExp)) +    geom_point() + expand_limits(y=0)
#expand_limits(y=0) function was added since the y-axis did not start from zero.
```



```{r}
#To summarize median gdp per capita within each continent with by_year_continent
by_year_continent <- gapminder %>% group_by(year, continent) %>% summarise(medianGdpPercap = median(gdp_cap))
head(by_year_continent, 10)
```


```{r}
#Plot a scatter plot median gdp per capita in each continent for by_year_continent (the expand_limits(y = 0) is because the y axis does not start from the origin).
ggplot(by_year_continent, aes(x = year, y = medianGdpPercap, color = continent)) +
  geom_point() + expand_limits(y = 0)
```
The median Gdp_cap increased over the years. Though, Africa had the lowest, while oceania had the highest.

```{r}
#Summarize the median gdp per capita and median life expectancy by continent in 2007
by_continent_2007 <- gapminder %>% filter(year == 2007) %>% group_by(continent) %>% summarise(medianGdpPercap = median(gdp_cap), medianLifeExp = median(life_exp))
by_continent_2007
```



```{r}
#Create the scatter plot of the by_continent_2007 data
ggplot(by_continent_2007, aes(x = medianGdpPercap, y = medianLifeExp, color = continent)) +    geom_point() + expand_limits(y = 0)
```



```{r}
#Creating Line Plots
#To summarize median gdp per capita by year
by_year_Gdp <- gapminder %>% group_by(year) %>% summarise(medianGdpPercap = median(gdp_cap))
head(by_year_Gdp, 10)
```


```{r}
#Create a line plot of the median gdp per capita for the year

ggplot(by_year_Gdp, aes(x = year, y = medianGdpPercap)) +    geom_line() + expand_limits(y = 0)
```



```{r}
#Summarize the median gpd per capita by year and continent
by_year_continent_Gdp <- gapminder %>% group_by(year, continent) %>% summarise(medianGdpPercap = median(gdp_cap))
head(by_year_continent_Gdp, 10)
```



```{r}
#To create a line plot for the by_year_continent_Gdp data
ggplot(by_year_continent_Gdp, aes(x = year, y = medianGdpPercap, color = continent)) + geom_line() + expand_limits(y = 0)
```



```{r}
#Creating Barplots
#Summarize the median gdp by continent in 1957
by_continent_1957 <- gapminder %>% filter(year == 1957) %>% group_by(continent) %>% summarise(medianGdpPercap = median(gdp_cap)) 
by_continent_1957
```

```{r}
#Plot the barchart of the by_continent_1957 data
ggplot(by_continent_1957, aes(x = continent, y = medianGdpPercap)) +  geom_col()
```



```{r}
#Creating Histogram
#Filter the dataset for the year 1957. Create a new column called population per million (pop_by_mil)
gapminder_1957_pop <- gapminder %>% filter(year == 1957) %>% mutate(pop_by_mil = population/1000000) 
head(gapminder_1957_pop, 10)
```



```{r}
#Create a histogram of the gapminder_1957_pop dataset
ggplot(gapminder_1957_pop, aes(x = pop_by_mil)) +    geom_histogram(bins = 50) + scale_x_log10()
```



```{r}
#Create the gapminder_1957 and filter for the year 1962
gapminder_1962 <- gapminder %>% filter(year == 1962)
head(gapminder_1962, 10)
```

```{r}
#create a box plot of the gapminder_1962 data
ggplot(gapminder_1962, aes(x = continent, y = gdp_cap)) +    geom_boxplot() + scale_y_log10()
```